---
title: "rpostgisLT: Use cases"
author: "Balázs Dukai, Mathieu Basille, David Bucklin, and Clément Calenge"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{rpostgisLT: Use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Motivation for the package and its relation to adehabitatLT

Recent technological progress allowed ecologists to obtain a huge
amount and diversity of animal movement data sets (usually from
wildlife collars/sensors) of increasing spatial and temporal
resolution and size, together with complex associated information
related to the environmental context, such as habitat types based on
remote sensing, population density, or weather. Such data often
require the use of an integrated database system, and a solution of
choice is the open-source database management system PostgreSQL, with
its extension PostGIS, that adds support for spatial data. Storing
spatial objects in a PostGIS-enabled database is particularly useful
for movement data, which can be very large, regularly updated, and
require cleaning and manipulation prior to being used in research.

On the other end of the process, the advancement of a movement ecology
theoretical framework led to an unprecedented development of new
analytical tools and methods, mostly available in the R statistical
environment.  The R package
[adehabitatLT](https://cran.r-project.org/package=adehabitatLT/)
is a collection of tools for the analysis of animal movements. In
particular, it builds on a dedicated class for animal movement data
(`ltraj` objects), which abstracts movement to a set of trajectories
and its geometrical descriptors.

**The package `rpostgisLT` focuses on streamlining the workflow for
biologists storing/processing movement data in PostGIS and analyzing
it in R, and aims at providing the tools to transparently benefit from
the power of the most advanced database and statistical systems
available for movement data. In particular, `rpostgisLT` provides full
integration with data type `ltraj` from `adehabitatLT` with
bi-directionnal conversion between PostGIS and R.**

Before getting started with `rpostgisLT`, we recommend reading the
[vignette of *adehabitatLT*](https://cran.r-project.org/web/packages/adehabitatLT/vignettes/adehabitatLT.pdf),
which clearly defines a trajectory and its elements. Let us however
repeat a few key concepts that are regularly used in this vignette. A
**trajectory** is a continuous curve described by an animal, person or
object when it moves. If a trajectory is sampled with e.g. a GPS
tracker, each measurement represents a **relocation**, while the
straight line segment that connects two successive relocations forms a
**step**.


## Use cases for the rpostgisLT package

**NOTE: infolocs of an `ltraj` are not implemented yet! Although infolocs have their
place in the database model, the `rpostgisLT` functions will ignore them.**

At the core, we are dealing with trajectories, which are loosely
defined objects, but essentially a sequence of points built into
successive steps. That's the starting point. Now, together PostgreSQL
and R provides tools to process, manage and analyse trajectories. Good
practice and the strengths of both tools would tend towards using
PostgreSQL for processing and management, and R for analysis, but
there is no strict border, and both allow some (or all) of the other
processes as well. In R, `adehabitatLT` actually allows for pretty
much everything using the `ltraj` class, which is already formally
defined. On top of this thus comes `rpostgisLT`: the essence of this
package is to allow for bidirectional transfer, at any times between
PostgreSQL and R, with no data loss or data alteration. This works by
establishing the corresponding structure of data in PostgreSQL
(`pgtraj` objects, see the
[vignette of the package](/web/packages/rpostgisLT/vignettes/data-model.html))
that stores (non destructively) all information from `ltraj` objects
in the database, and allows for the use of PostgreSQL tools (notably
PostGIS) on the data.


### Initialization

A typical session will start by initializing the connection to the
database, using PostgreSQL credentials through the `RPostgreSQL`
package (note that loading `rpostgisLT` automatically loads
`RPostgreSQL` too):

```{r eval=FALSE}
library(rpostgisLT)
con <- dbConnect("PostgreSQL", dbname = <dbname>, host = <host>, user = <user>, password = <password>)
```

The next step will be to check whether the intended pgtraj schema is
ready to use with the function `pgTrajSchema` (note that by default,
the function checks and/or create the schema "traj", which can be
changed with the `name` argument):

```{r eval=FALSE}
pgTrajSchema(con)
```

If it is successful, the function should return `TRUE`, together with
a message like:

```
The pgtraj schema 'traj' was successfully created in the database.
```

Or in the case of an existing schema:

```
The schema 'traj' already exists in the database, and is a valid pgtraj schema.
```



***********************************************



### Basic transfer

The most basic feature demonstrates a simple transfer from R to
PostgreSQL and back to R: the resulting object should equal the
original.

```{r eval=FALSE}
data(ibexraw)
ibexraw
is.regular(ibexraw)
## FALSE

## Note that there is an issue with the time zone. In 'ibexraw', the
## time zone is not set:
attr(ld(ibexraw)$date, "tzone")
## This means that it is assumed to be UTC, and is thus converted to
## local time zone on display (EDT2EST for me):
head(ld(ibexraw)$date)                 # Note that the first timestamp
                                       # should be '2003-06-01
                                       # 00:00:56'
## We need to fix that upfront:
ibex <- ld(ibexraw)
attr(ibex$date, "tzone") <- "Europe/Paris"
ibex <- dl(ibex)

ltraj2pgtraj(ibex)                       # Default should be in schema
                                             # 'traj' and use ltraj name
                                         # ('ibex') as pgtraj name.
ibexTest <- pgtraj2ltraj(con, "ibex")    # Default should look into
                                        # 'traj' schema.
all.equal(ibex, ibexTest)
## TRUE
```

Note that changes were implemented to the `ltraj` data structure in `adehabitatLT` v0.3.21:  1) row names are character strings; 2) there is an additional attribute *proj4string* in an ltraj that stores the projection reference. The adehabitatLT package must be updated to that version in order to install `rpostgisLT`; however, old `ltraj`s that were created with a previous version of `adehabitatLT` should still work with `rpostgisLT`, and can be manually updated to include a proj4string using a valid `CRS` object as follows:

```{r eval=FALSE}
attr(ltraj, "proj4string") <- CRS("+proj=longlat +datum=WGS84")
```

Now there are many ways to alter a trajectory (in R or PostgreSQL), but each of them should run smoothly on either side while being transferrable at any time to the other side. Each of these modifications should end up with the same test, i.e. that `all.equal` between the original R object and the one that has been stored in PostgreSQL and retrieved in R returns `TRUE`.


### Missing steps [seq, t, dt]

An `ltraj` can include NAs in their sequence (i.e. missing relocations), but still provide a record of them with their timestamp but no coordinates (as is the case with the example dataset `puechcirc`). On the other hand, `ibexraw` only provides a record when coordinates are available. We can add missing relocations using `setNA`:

```{r eval=FALSE}
refda <- strptime("2003-06-01 00:00", "%Y-%m-%d %H:%M", tz = "Europe/Paris")
(ibex <- setNA(ibex, refda, 4, units = "hour"))
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
## TRUE
```

### Regularize [t, dt]

The next logical step is to regularize the trajectory, by "rounding" timestamps to their expected values using `sett0'`:

```{r eval=FALSE}
(ibex <- sett0(ibex, refda, 4, units = "hour"))
ibex.bkp <- ibex                        # At this stage, 'ibex' is our
                                        # reference data
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
## TRUE
```

### Interpolate [seq, geom, t, dt]

Two types of interpolation can be computed: in space, i.e. rebuilding a trajectory based on a given step length; and in time, i.e. linearly interpolate missing data. 

```{r eval=FALSE}
## 1. In space
summary(ld(ibex)$dist)
(ibex <- redisltraj(ibex, 400))
ibex <- removeinfo(ibex)
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
## TRUE

## 2. In time
ibex <- ibex.ref
(ibex <- redisltraj(na.omit(ibex), 14400, type = "time"))
ibex <- removeinfo(ibex)
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
## TRUE
```

### Subset the trajectory [seq, (geom), (dt)]

In practice, there are two ways to subset a trajectory: by querying its parameters (or infolocs) on a specific condition, or by subsampling the trajectory at regular intervals.

```{r eval=FALSE}
## 1. Subset on given parameters
ibex <- ibex.ref
## We work on the data frame from the trajectory, which we subset, and
## then rebuild the ltraj without recomputing trajectory parameters;
## this is essentially what 'hab::subset' does.
## Note that the steps are not continuous any more.
ibex <- ld(ibex)
ibex <- droplevels(ibex[ibex$dist < 400 & !is.na(ibex$dist), ])
dlfast <- function(x) {
    trajnam <- c("x", "y", "date", "dx", "dy", "dist", "dt",
        "R2n", "abs.angle", "rel.angle")
    idd <- tapply(as.character(x$id), x$burst, unique)
    traj <- split(x[, names(x) %in% trajnam], x$burst)
    names(traj) <- NULL
    class(traj) <- c("ltraj", "list")
    attr(traj, "typeII") <- TRUE
    attr(traj, "regular") <- is.regular(traj)
    for (i in (1:length(traj))) {
        attr(traj[[i]], "id") <- as.character(idd[i])
        attr(traj[[i]], "burst") <- names(idd[i])
    }
    return(traj)
}
ibex <- dlfast(ibex)
head(ibex[[1]])
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)

## 2. Subsample on the temporal sequence
ibex <- ibex.ref
(ibex <- subsample(ibex, 14400*2))
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
```

### Cut, bind bursts [burst]

Sometimes, it is useful to cut a trajectory into sub-bursts, based on a given condition assessed on the trajectory parameter. For instance, we may want to cut into different bursts when steps are too large:

```{r eval=FALSE}
## 1. Cut if there is a step greater than 3000 m
ibex <- ibex.ref
(ibex <- cutltraj(ibex, "dist > 3000"))
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
```

The opposite process is to bind bursts from a unique individual into a single burst:

```{r eval=FALSE}
## 2. Bind back by individual:
(ibex <- bindltraj(ibex))
ibex <- removeinfo(ibex)
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
```

### Combine trajectories or bursts [burst, traj]

The structure of a ltraj allows to combine different ltraj objects (or selection of bursts); in other word, a ltraj is a collection of bursts (technically a `list`), which can be manipulated just as any R `list`:

```{r eval=FALSE}
ibex <- ibex.ref
ibex2 <- ibex
burst(ibex2) <- paste(burst(ibex2), "2", sep = "-")
(ibex <- c(ibex, ibex2)[order(id(c(ibex, ibex2)))])
ltraj2pgtraj(con, ibex, overwrite = TRUE)
ibexTest <- pgtraj2ltraj(con, "ibex")
all.equal(ibex, ibexTest)
```



 
